Dissociation between vertebrae number and body size in Australian blindsnakes

### 2024-07-18 manuscript.Rmd
update before sending to Emma on 2024-07-18

* Axial evolution an important aspect of the vertebrate body plan evolution – facilitates diversity and influences locomotion 
  <!-- Note the papers  that are about axial number relating to locomotion -->
@galisFast2014 - mammals
@vandammeSpeed2002 - lacertids
@yamadaAdaptive2009;@forkerBendy2021 - gobies 
@molnarMorphological2023 - chameleons

@@

* Elongate animals have special properties of vertebral number known as pleomerism, where body size and number of segments in vertebrae are correlated [@lindseyPleomerism1975]. 
Pleomerism has been observed in eels and fish [@wardEvolution2007;@wardDifferential2014;@reeceEvolutionary2013;], in squamates [@pollyTesting2001;@headDissociation2007;@headEvolution2015;@bergmannVertebral2012;@lindellEvolution1994;@cernanskyVertebral2019], in salamanders and caecilians [@bonettEvidence2017;@renousBody1989;@arntzenVertebral2015]

@@

Scolecophidians, often referred to in English as "blindsnakes", are characterised by their fossorial lifestyle, poor visual system, and highly specialised phenotypes adapted for a strictly fossorial existence [@mirallesMolecular2018; @strongDeconstructing2021; @gowerEyetranscriptome2021]. Scolecophidians comprise five paraphyletic families, including the "core scolecophidians" — a monophyletic clade consisting of Typhlopidae (425 spp), Leptotyphlopidae (143 spp), Gerrhopilidae (29 spp), and Xenotyphlopidae (1 spp) — and the paraphyletic Anomalepididae (20 spp), which is sister to all other "typical snakes" (Alethinophidia) [@mirallesMolecular2018; @burbrinkInterrogating2020; @uetzQuarter2021].

Although all blindsnakes are fossorial, body size and shape vary significantly among species (Tiatragul et al. unpublished; @shineNatural1990;@webbPreysize1993;@webbDietary2001;@ehmannFamily1993). Leptophlopids are generally thought to be more narrow-and-long compared to typhlopids, which are generally wide-and-robust for a given body length [@herrelBurrowing2021]. These differences in body size and proportions determine how much push force the snakes can generate when burrowing. Their morphological differences led to the idea that leptyphlopids generally use existing burrows while typhlopids create their own burrow.

@@

All statistical analyses were performed in R Statistical Environment. To test for evolutionary correlations between the number of vertebrae and body size and proportions, we used phylogenetic generalised least squares (PGLS) regression under a model of Brownian Motion implemented in the R package *`geomorph`* v.4.0.6 [@adamsGeomorph2023]. First, we tested for pleomerism [@lindseyPleomerism1975] across 37 species of *Anilios* blindsnakes. We did this by testing whether the total number of vertebrae positively correlate with body size when regressing the maximum number of total vertebrae against maximum total body length from Tiatragul et al. (unpublished). Next, we test whether the vertebrae to body length ratio is related to the mean vertebrae number and mean aspect ratio, defined by total body length / body width, also from Tiatragul et al. (unpublished) data. Finally, we tested if variation in vertebrae ratio and vertebrae number can be explained by mean annual temperature and soil compactness. In all PGLS analyses, statistical significance was assessed by permutation, using 1000 iterations.

To visualise the relationship between vertebrae number and intracolumnar vertebral length, for each of the representative species, we plotted the mean vertebrae size (from three repeated measurements) against the vertebra number. Each vertebrae profile was overlaid with a fitted 4° polynomial curve.

@@

For the Australian blindsnakes, we found a weakly positive but non-statistically significant relationship between maximum number of total vertebrae and maximum SVL (PGLS, slope=0.28, intercept=256, R^2^=0.02, F=0.74, Z=0.30, P=0.41), indicating an absence of pleomerism in Australian blindsnakes (Fig. \@ref(fig:fig-tbl-vert)). This suggests that maximum length is achieved by elongating each vertebra rather than by adding additional vertebrae, while most species have many more vertebrae than expected for their size. For example, *A. grypus* "West" is an intermediate length (relative to other species) and has the most vertebrae, while *A. torresianus*, which is the longest, has relatively few vertebrae.

### Relationship between vertebrae ratio, number, and body shape
We found that vertebrae ratio is positively correlated with aspect ratio (PGLS, slope=-0.0048, intercept=0.73, R^2^=0.17, F=7.09, Z=2.20, P=0.011), indicating that species that are more narrow-and-long usually have relatively more vertebrae for their length than species that are wide-and-robust (Fig. \@ref(fig:fig-vert-aspect-ratios) and \@ref(fig:fig-vertebral-profile)). We also found that the mean number of vertebrae positively correlated with aspect ratio (PGLS, slope=-1.8, intercept=120.95, R^2^=0.69, F=79.39, Z=5.30, P<0.001).

@@

knitr::include_graphics(paste0(WD, '/assets/figures/tbl-vert.jpg'))
```

<!-- Fig: Pleomerism and ratio correlations -->
```{r fig-vert-aspect-ratios, fig.cap= 'Species mean aspect ratios against (A) mean vertebrae ratio and (B) mean number of vertebrae of Australian blindsnakes. Vertebrae ratio is calculated by total vertebrae/total length, and aspect ratio is total body length/midbody diameter. Species with large aspect ratios are narrow-and-long while species with low aspect ratios robust-and-wide.', out.width="124mm", out.height= "165mm"}

knitr::include_graphics(paste0(WD, '/assets/figures/vert-aspect.jpg'))
```

<!-- Fig: Vertebral profile -->
```{r fig-vertebral-profile, fig.cap= 'Relative vertebral length plotted against vertebra number for three example species with fitted 4° polynomial curves overlaid.', out.width="140mm", out.height= "197mm"}

knitr::include_graphics(paste0(WD, '/assets/figures/vertebral-profile.jpg'))
```

@@

<!-- Fig: Ratio and ecological variables -->
```{r fig-vert-ecology, fig.cap= 'Relationship between mean vertebrae ratios and ecological variables. (A) Mean vertebrae against species mean annual temperature. (B) Mean vertebrae ratio against species max soil bulk density. Mean vertebrae ratio is calculated by mean total vertebrae / mean total length for each species. Species mean annual temperature and max soil bulk density values are calculated based on occurrence data. Trend lines represent slopes from PGLS', out.width="165mm", out.height= "117mm"}

knitr::include_graphics(paste0(WD, '/assets/figures/vert-ecology.jpg'))
```

@@

@@

>>>>

* Axial evolution an important aspect of the vertebrate body plan evolution – facilitates diversity and influences locomotion 
  <!-- Note the papers  that are about axial number relating to locomotion -->
in mammals [@galisFast2014 ], lacertids [@vandammeSpeed2002], gobies [@yamadaAdaptive2009;@forkerBendy2021], and chameloens [@molnarMorphological2023].

@@

* Elongate animals have special properties of vertebral number known as pleomerism, where body size and number of segments in vertebrae are correlated [@lindseyPleomerism1975]. 
Pleomerism has been observed in eels and fish [@wardEvolution2007;@wardDifferential2014;@reeceEvolutionary2013], in squamates [@pollyTesting2001;@headDissociation2007;@headEvolution2015;@bergmannVertebral2012;@lindellEvolution1994;@cernanskyVertebral2019], in salamanders and caecilians [@bonettEvidence2017;@renousBody1989;@arntzenVertebral2015].

@@

Scolecophidians, often referred to in English as "blindsnakes", are characterised by their fossorial lifestyle, poor visual system, and highly specialised phenotypes adapted for a strictly fossorial existence [@mirallesMolecular2018;@strongDeconstructing2021;@gowerEyetranscriptome2021]. Scolecophidians comprise five paraphyletic families, including the "core scolecophidians" — a monophyletic clade consisting of Typhlopidae (425 spp), Leptotyphlopidae (143 spp), Gerrhopilidae (29 spp), and Xenotyphlopidae (1 spp) — and the paraphyletic Anomalepididae (20 spp), which is sister to all other "typical snakes" (Alethinophidia) [@mirallesMolecular2018; @burbrinkInterrogating2020; @uetzQuarter2021].

Although all blindsnakes are fossorial, body size and shape vary significantly among species [@shineNatural1990;@webbPreysize1993;@webbDietary2001;@ehmannFamily1993]. Leptophlopids are generally thought to be more narrow-and-long compared to typhlopids, which are generally wide-and-robust for a given body length [@herrelBurrowing2021]. These differences in body size and proportions determine how much push force the snakes can generate when burrowing. Their morphological differences led to the idea that leptyphlopids generally use existing burrows while typhlopids create their own burrow.

@@

All statistical analyses were performed in R Statistical Environment. To test for evolutionary correlations between the number of vertebrae and body size and proportions, we used phylogenetic generalised least squares (PGLS) regression under a model of Brownian Motion implemented in the R package *`geomorph`* v.4.0.6 [@adamsGeomorph2023]. First, we tested for pleomerism [@lindseyPleomerism1975] across 37 species of *Anilios* blindsnakes. We did this by testing whether the total number of vertebrae positively correlate with body size when regressing the maximum number of total vertebrae against maximum total body length from Tiatragul et al. (unpublished). Next, we test whether the vertebrae to body length ratio is related to the mean vertebrae number and mean aspect ratio, defined by total body length / body width, also from Tiatragul et al. (unpublished) data. Finally, we tested if variation in vertebrae ratio and vertebrae number can be explained by mean annual temperature and soil compactness. In all PGLS analyses, statistical significance was assessed by permutation, using 1000 iterations. 
To visualise the relationship between vertebrae number and intracolumnar vertebral length, for each of the representative species, we plotted the mean vertebrae size (from three repeated measurements) against the vertebra number. Each vertebrae profile was overlaid with a fitted 4° polynomial curve.

@@

For the Australian blindsnakes, we found a weakly positive but non-statistically significant relationship between maximum number of total vertebrae and maximum SVL (PGLS, slope=0.28, intercept=256, R^2^=0.02, F=0.74, Z=0.30, P=0.41), indicating an absence of pleomerism in Australian blindsnakes (Fig. \@ref(fig:fig-tbl-vert) and **Fig S1**). This suggests that maximum length is achieved by elongating each vertebra rather than by adding additional vertebrae, while most species have many more vertebrae than expected for their size. For example, *A. grypus* "West" is an intermediate length (relative to other species) and has the most vertebrae, while *A. torresianus*, which is the longest, has relatively few vertebrae.

### Relationship between vertebrae ratio, number, and body shape
We found that vertebrae ratio is positively correlated with aspect ratio (PGLS, slope=-0.0048, intercept=0.73, R^2^=0.17, F=7.09, Z=2.20, P=0.011), indicating that species that are more narrow-and-long usually have relatively more vertebrae for their length than species that are wide-and-robust (Fig. \@ref(fig:fig-vert-aspect-ratios)). We also found that the mean number of vertebrae positively correlated with aspect ratio (PGLS, slope=-1.8, intercept=120.95, R^2^=0.69, F=79.39, Z=5.30, P<0.001).

@@

knitr::include_graphics(paste0(WD, '/assets/figures/phylomorpho-tbl-vert.jpg'))
```

<!-- Fig: Pleomerism and ratio correlations -->
```{r fig-vert-aspect-ratios, fig.cap= 'Species mean aspect ratios against (A) mean vertebrae ratio and (B) mean number of vertebrae of Australian blindsnakes. Vertebrae ratio is calculated by total vertebrae/total length, and aspect ratio is total body length/midbody diameter. (C) Vertebral profile shape of three species representing different vertebrae ratios, and are depicted as a 4° polynomial curve with positive inflection. Species with large aspect ratios are narrow-and-long while species with low aspect ratios robust-and-wide.', out.width="124mm", out.height= "165mm"}

knitr::include_graphics(paste0(WD, '/assets/figures/phylomorpho-vert-aspect-total.jpg'))
# knitr::include_graphics(paste0(WD, '/assets/figures/vert-aspect.jpg')) -->
```

<!-- Fig: Vertebral profile -->
<!-- ```{r fig-vertebral-profile, fig.cap= 'Relative vertebral length plotted against vertebra number for three example species with fitted 4° polynomial curves overlaid.', out.width="140mm", out.height= "197mm"}

knitr::include_graphics(paste0(WD, '/assets/figures/vertebral-profile.jpg'))
``` -->

@@

<!-- Fig: Ratio and ecological variables -->
```{r fig-vert-ecology, fig.cap= 'Phylomorphospaces showing relationship between mean vertebrae ratios and mean vertebrae number against ecological variables. Mean vertebrae ratio against (A) species mean annual temperature and (B) species max soil bulk density. Mean vertebrae ratio is calculated by mean total vertebrae / mean total length for each species. Mean vertebrae number against (C) species mean annual temperature and (D) species max soil bulk density. Species mean annual temperature and max soil bulk density values are calculated based on occurrence data. Trend lines represent slopes from PGLS', out.width="165mm", out.height= "117mm"}

knitr::include_graphics(paste0(WD, '/assets/figures/phylomorpho-vert-ecology.jpg'))
```

@@

Discuss amphisbaenians [@gansAmphisbaenia1978] and caecillians [@renousBody1989].

@@

REGIONALISATION Need to mention, plus the heart position aspect; how that’s not been studied here, but it helps to understand regionalisation, which is an important next step in understanding the lack of pleomerism. Since the lack of pleomerism in *Hydrophis* sea snakes is due to their elongate forebodies. Tamika’s paper on colubrids also shows the various ways elongation occurs, using different regions. Hampton 2023 paper on snake regionalisation in greater detail
